The Eclipse Integrated Computational Environment

### 2017-01-09 src/content.tex
First full draft of the article.
Merged manually from github.com/jayjaybillings/thesis. Includes edits
from Sam Crawford and the most recent copy of the article and
bibliography.

Signed-off-by: Jay Jay Billings <billingsjj@ornl.gov>

High performance modeling and simulation software is hard to use. Much
of the challenge comes from the inherent nature of the science under
consideration, but perhaps just as much of the problem comes from
focusing very heavily on raw compute performance. Focusing on
performance is understandable, given the field, but it leaves the rest
of us to figure out how to actually use the software in a way that
scales for a much larger set of users, many of whom may be novices or
migrate between software packages regularly.

** REMOVE or merge with text below!**

In previous work, Billings et. al., discovered through requirements
gathering interviews that many of the difficulties using high
performance modeling and simulation software fall broadly into five
distinct categories, \cite{billings_cbhpc}. This includes input
generation and preprocessing (also called ``model setup''); job
execution and monitoring; postprocessing, visualization and data
analysis; data management; and customizing the software. There are many
tools that address these problems individually, but the same work found
that the excess number and specialization of these tools also contribute
to the learning curve.

Efforts to address these five issues typically fall in with general
purpose scientific workflow tools like Kepler, \cite{kepler}, or are
reduced to myopic tools that satisfy some set of requirements for a
single piece of software or platform. That is, the proposed solution to
this problem is often to shoe-horn it into existing workflow tools that
are so general that they focus on nothing in particular or to ignore the
general problem entirely and deploy a completely tailored solution for
the given application. These are opposing extremes, but a
middle-of-the-road solution is also possible. A workflow engine could be
developed that limits its scope to High-Performance Computing (HPC) and
to the set of possible workflows that come from the previously mentioned
five activities. A rich enough Application Programming Interface (API)
could be exposed so that highly customized solutions could still be made
based on this limited workflow engine with only a relatively minor
amount of additional development required.

It is not clear that one of these solutions is better than the others.
Practical requirements will ultimately dictate which way projects go.
This work considers a middle ground solution and presents the Eclipse
Integrated Computational Environment (ICE) as proof that it is possible
to create such a system. Specifically, we show * an architecture for
such a workflow system that satisfies the model of workflows presented
below in an extensible way. * that such a system can be cross-platform
for workstations and simultaneously deployable on the web. * that smart
design decisions enable not only authoring code for simulation software,
but also make it possible for the system to extend itself, thereby
enabling heavy customization.

\textbf{FIXME! Needs to transition smoothly!}

@@

Computational scientists perform a variety of tasks with respect to
modeling and simulation that can be abstracted into a lightweight
theoretical framework broadly based on five high-level
\emph{activities}: creating input, executing jobs, analyzing results,
managing data, and modifying code. The same computational scientist
would most likely find these activities difficult for all codes with
which they lack experience, whereas with their own codes or those with
which they are most familiar these tasks may be so easy that they are
taken for granted. Any particular combination of these activities across
one or more scientific software packages or codes results in a unique
\emph{workflow}. Such a workflow is normally, but not always, requested
by a human user and orchestrated by a \emph{Workflow Management System
(WMS)}.

The most obvious workflow for any individual simulation code or
collection of codes is to string the activities together: the user's
workflow is to create the input, launch the job, perform some analysis
and manage their data, possibly modifying their code in the process.
There are many other combinations though, such as re-running jobs with
conditions or modifications or analyzing someone else's data. (See
footnote 1.)

\textbf{Creating input} is the process of describing the physical model
or state of a system that will be simulated. This could include creating
an input file or file(s), but could also include external calls to a
process to configure a running program. In most situations a
computational scientist will modify existing input or create new input
from a template. ``Input'' generally includes runtime parameters for the
simulation framework (tolerances, etc.); configuration options (data
locations, output locations, module configurations, etc.); material
properties for the materials that will be simulated; and a
discretization of the simulation space (mesh, grid, particle
distribution, etc.). The collection of all required input can be quite
large and may go by many names such as ``input set,'' ``input package,''
``problem,'' or, simply, ``input.'' Often the set of input files will be
described in a ``main'' input file that acts as a kind of manifest,
describing and providing links to all necessary information for the
given problem.

In this work, it should be assumed unless otherwise noted that ``input''
refers to the entire set of input, not a single file.

\textbf{Executing jobs} or ``running the workflow'' in this context is
the process of performing some calculation based on the input with a
simulation code or framework. Modeling and simulation codes are normally
run locally for small jobs or development. Large simulations typically
require equally large hardware resources that are utilized remotely
through Secure Shell (SSH) connections or similar protocols. Remote
execution requires moving the input in advance of the execution and
copying or moving the output to the user's machine, if possible and
desirable. (In many cases the output cannot be moved because it is too
large.)

Local and remote jobs are often monitored in one or more ways to
ascertain the status of job. This can include simply checking whether or
not execution has finished or monitoring the output of individual
quantities to examine the calculation state. The latter is often used to
detect calculation errors that will result in incorrect results. If such
results are found, the job is typically cancelled (``killed'') to save
compute resources and re-run later.

In this work, it should be assumed unless otherwise noted that
``executing a job'' includes monitoring that job in one or more ways,
possibly including real-time updates to visualizations.

\textbf{Analyzing results} includes executing special jobs to transform
data in one or more prescribed ways and producing artifacts with
scientific significance from the transformed data. This may include, for
example, post-processing results and visualizing the new data with
dedicated visualization tools. For many types of scientific computing
this includes viewing the results of a simulation on a mesh or grid and
extracting publication quality images or movies from that data. Other
cases may include analyzing results in preparation for follow
simulations or performing feature extraction, classification or
activities for machine learning and data mining.

While this has many similarities to executing a job, it is distinctly
different because the activity changes focus to satisfy the needs of a
human operator. Simple data reduction where the exact reduction is known
certainly qualifies as executing a job, but analysis of modeling and
simulation results is far from simple data reduction. It is generally a
far more interactive process for computational scientists.

\textbf{Managing data} includes moving, copying, storing, sharing or
otherwise interacting with data for or from simulations. This activity
is the most pervasive of all the activities because each of the others
requires interacting with data in some way. In many cases though, data
is still managed its own purposes, without performing a simulation,
generating new input, or analysing results. Examples include archiving
data, packaging data for publications, and updating values manually
(often in light of new information from publications).

\textbf{Modifying code} is not typically considered a part of a
scientific computing workflow. However, modeling and simulation use
cases often require users to explicitly modify code before execution,
such as with the computational fluid dynamics code Nek5000,
{[}Nek5000{]}; or to issue special re-build instructions. Many
computational scientists consider ``their workflow'' to be re-running
software after modifications for purely exploratory purposes. This may
be required if the model the author is manipulating can not be
configured as part of the input directly but can be easily manipulated
by hand.

@@

This model of workflows differs significantly from those of similar
efforts in workflow science because it defines workflows in terms of
activities. Others in the literature define a workflow as a collection
of computing processes. For example, Yu and Buyya define grid workflows
as ``a collection of tasks that are processed on distributed resources
in a wel-defined order to accomplish a specific goal,'' {[}Yu, 2005{]}.
Others such as Pizzi et al. subscribe to similar definitions, {[}Pizzi,
2015{]}. This ``process'' view is acceptable where the workflow is
static and does not require additional human input or ``human in the
loop'' behavior after the all the initial human input is provided. That
is, a ``process-oriented'' definition is acceptable where all human
input is provided in advance. However, workflows within ICE are fully
interactive with regular call-backs to humans. It is simpler to discuss
``activities'' than it is to create a distinction between ``human
processes'' and ``computer processes.'' Focusing on ``activities'' over
processes (human or computer) also has the benefit of removing concrete
elements such as hardware properties or software details that distract
from details of workflows and WMSes per se. That is, considerations such
as memory usage and raw performance are important, but questions about
the abstract workflow or what the WMS should do are far more important.

@@

Workflows and tasks in ICE are not explicitly treated as trees -
Directed Acyclic Graphs, (DAGS) - as is common with grid workflow tools,
{[}Yu, 2005{]}. Instead, ICE's design is inspired by Representational
State Transfer (REST), {[}Fielding, 2000{]}.

@@

Individual components of workflows, i.e. - workflow ``tasks'' or
``nodes'' - are either encoded directly in the workflow's subclass of
Item or provided as \emph{Actions} that are dynamically registered with
an \emph{Action Factory} and obtained at runtime. Table 1. describes the
differences between Items, Actions, and Forms. (See footnote 2.)

@@

\begin{longtable}[c]{@{}lll@{}}
\toprule
Class & Class Description & Object Description\tabularnewline
\midrule
\endhead
Item & Java class with code to execute an abstract workflow. Provides a
Form. & Concrete workflow executor.\tabularnewline
Form & Description and template of the data needed for the Item to
process the workflow. & User-modified workflow data.\tabularnewline
Action & Java class for executing a specific task in the workflow. Used
by Items. & Concrete workflow task executor.\tabularnewline
\bottomrule
\end{longtable}

@@

ICE handles persistence using a \emph{persistence provider}. The only
default persistence provider uses JAX-RS, {[}Burke, 2009{]}, to write
Forms to XML. In principle other persistence providers could replace the
XML-based provider since it is registered as a dynamic service, and in
the past a JAX-P based provider has been used.

@@

ICE is an Eclipse Rich Client Platform (RCP) application, {[}McAffer 2,
2010{]}. It has a plugin architecture based on Equinox, the reference
implementation of the Open Service Gateway Initiative (OSGi) framework
specification, {[}McAffer, 2010{]}. It uses over 1200 additional
packages from the Eclipse ecosystem to provide services such as language
support and visualization to users. Each unique element of ICE described
in this work, including the client, core, all Items, data structures and
many others are provided as plugins to Equinox. Most plugins are
dynamically managed and provided as services that can be obtained from a
service registry. All file I/O in ICE, with only a few exceptions,
interacts with the RCP's Resources plugin. This includes remote
resources that are managed with the Eclipse Parallel Tools Platform,
{[}Tibbitts, 2009{]}.

@@

There are many situations where graphically configuring workflows is
unacceptable, such as when a very large number of workflows will be
executed or the type of information required is very fine-grained. In
these cases, it is often necessary to provide scripts to the workflow
management system. ICE includes the Eclipse Advanced Scripting
Environment, {[}Pontesegger, 2017{]}, (EASE), for scripting because it
provides a native way to script Eclipse RCP projects in Javascript,
Jython, and Python. This also makes it possible to extend the
environment by adding Items in these languages.

@@

Outside of the development team, ICE is commonly deployed as a
sophisticated user environment for computational science projects (see
``Impact'' below) and as a visualization tool. ICE originally a
significant amount of visualization support, but at the request of users
in the community that support was ``spun off'' in early 2016 as a
separate visualization project, the Eclipse Advanced Visualization
Project, EAVP, {[}EAVP, 2009{]}.

@@

Pannala et al. developed a Virtual Integrated Battery Environment (VIBE)
as part of their research into safety and performance characteristics of
lithium ion batteries, {[}Pannala, 2015{]}. VIBE provides ICE to users
as part of its distribution. New workflows were added to ICE to create
multiple types of input, configure the simulation software, and launch
simulations of virtual batteries. Interactive 3D visualizations of the
results were embedded in the launcher so that users would quickly find
their results. Figure 1 shows the results from a temperature
distribution of a prismatic cell battery simulated during discharge.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/vibe_20151016.png}
\caption{vibe}
\end{figure}

@@

The MOOSE Framework is a powerful, easy to use multiphysics framework
developed at the Idaho National Laboratory, {[}Gaston, 2009{]}. ICE
provides workflow tools for MOOSE as well as specialized class
generation utilities for developing custom MOOSE kernels. Many of the
MOOSE tools in ICE were developed closely with the MOOSE team to closely
reproduce various aspects of MOOSE's user interface, Peacock. Figure 2
shows an example of the ICE workbench for a simple structural mechanics
problem solved with the MOOSE framework, {[}McCaskey, 2016{]}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/ice-moose.png}
\caption{moose}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/ice-bjm.png}
\caption{bjm}
\end{figure}

@@

ICE includes a small utility for simulating neutron reflectivity and
comparing the results to data, {[}Billings, 2015{]}. This was developed
in collaboration with colleagues at the Spallation Neutron Source to
replace an older utility written in Visual Basic and distributed via
Excel macros. The newer utility in ICE is capable of processing a single
workflow and is shown in figure 4.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/reflectivity-screenshot.png}
\caption{reflectivity}
\end{figure}

@@

As quantum computing grows, so to does the need for sophisticated
software that can utilize quantum hardware or perform calculations on
simulated quantum hardware. Humble et al. created a simulator for
adiabatic quantum computers where workflows were added to ICE to support
interaction with the simulator, but also to process large sets of
quadratic binary optimimzation problems, {[}Humble, 2014{]}. Figure 5
shows the workbench for this project.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/jaded.png}
\caption{qc}
\end{figure}

ICE also supports other quantum computing projects, including the
eXtreme-scale ACCelerator (XACC) programming framework,
{[}ORNL-QCI/XACC, 2016{]}. XACC is a programming framework for
extreme-scale, post-exascale accelerator architectures that integrates
alongside existing conventional applications.

@@

There are many examples of ICE's role in modeling and simulation
projects for nuclear energy, but readers are referred to the work in
{[}Billings, 2015{]} for an extreme example of the level of
customization that is possible in ICE. Support for the ``Reactor
Analyzer'' was dropped in ICE 2.1.8, but it demonstrated ICE's ability
to integrate many different nuclear energy tools for complex analysis.

@@

One pressing area of interest and impact is that of interoperability
between workflow systems. There is significant prior art in combining
large workflow management systems, such as that Mandal et al.,
{[}Mandal, 2007{]}, but the end goal of gaining the greatest advantage
by using the best capabilities from multiple systems remains elusive.
ICE's unique perspective on workflows and its well-defined API make it
possible to integrate multiple systems in a straightforward way. This
allows it to connect to other workflow environments, such as Triquetrum,
quite easily, {[}Brooks 1, 2015{]}. Triquetrum, (like Kepler in
{[}Mandal, 2007{]}), is a Ptolemy-based workflow engine, {[}Brooks 2,
2015{]}. Several of the authors of this paper are using it to
investigate these issues.

@@

A new cloud based development tool based on ICE is under development by
RNET LLC out of Dayton, Ohio in response to an SBIR award from the U.S.
Department of Energy. This web based version of ICE will continue ICE's
support for nuclear energy and integrate with cloud services such as
Amazon Web Services and Oak Ridge National Laboratory's Compute and Data
Environment for Science (CADES). Additionally, while ICE has not led to
the development of ``spin-off'' companies, ICE source code has been used
in two spin-off projects: EAVP (mentioned earlier) for advanced
visualizations and the Eclipse January project for scientific data
structures, {[}Graham, 2016{]}. ICE was also one of the founding
projects of the Eclipse Science Working Group.

@@

The quintessential resource for information on ICE is the projects web
page, {[}ICE, 2016{]}. The ``Resources'' menu includes links to detailed
tutorials and documentation of multiple types. Examples of how to create
new workflow Items are available at
https://github.com/eclipse/ice/tree/master/org.eclipse.ice.demo.
Examples of how to use the scripting engine (in this case for neutron
reflectivity) are available at
https://github.com/eclipse/ice/tree/master/examples. ICE includes an
extensive test suite of unit, integration, and UI tests and these are
also excellent examples of how to work with the platform.

@@

--------- \textbar{}
:-----------------------------------------------------------------:
\textbar{} :-------------------------------------------------------- :
\textbar{}\\C1 \textbar{} Current Code Version \textbar{} `next'
\textbar{}\\C2 \textbar{} Permanent link to code/repository used for
this code version \textbar{} https://github.com/eclipse/ice/tree/next
\textbar{}\\C3 \textbar{} Legal Code License \textbar{} Eclipse Public
License 1.0 \textbar{}\\C4 \textbar{} Code versioning system use
\textbar{} Git \textbar{}\\C5 \textbar{} Software code languages, tools,
and services used \textbar{} Java, OSGi, Eclipse Rich Client Platform,
and Maven \textbar{}\\C6 \textbar{} Compilation requirements, operating
environments \& dependencies \textbar{} Java 1.8 or greater, Maven, and
an internet connection for dependencies \textbar{}\\C7 \textbar{} If
available Link to developer documentation/manual \textbar{}
https://wiki.eclipse.org/ICE \textbar{}\\C8 \textbar{} Support email for
questions \textbar{} ice-dev@eclipse.org \textbar{}

@@

--------- \textbar{}
:-----------------------------------------------------------------:
\textbar{} :-------------------------------------------------------- :
\textbar{}\\S1 \textbar{} Current Code Version \textbar{} 2.2.1
\textbar{}\\S2 \textbar{} Permanent link to executables of this version
\textbar{}
http://www.eclipse.org/downloads/download.php?file=/ice/builds/2.2.1/
\textbar{}\\S3 \textbar{} Legal Software License \textbar{} Eclipse
Public License 1.0 \textbar{}\\S4 \textbar{} Computing
platforms/Operating Systems \textbar{} Windows (32/64-bit), Mac OS/X,
Linux (32/64-bit) \textbar{}\\S5 \textbar{} Installation requirements \&
dependencies \textbar{} Java 1.8 or greater \textbar{}\\S6 \textbar{} If
available, link to user manual - if formally published include a
reference to the publication in the reference list \textbar{}
https://wiki.eclipse.org/ICE \textbar{}\\S7 \textbar{} Support email for
questions \textbar{} ice-users@eclipse.org \textbar{}

\section{References}\label{references}

Yu, Jia, and Rajkumar Buyya. ``A Taxonomy of Workflow Management Systems
for Grid Computing.'' Journal of Grid Computing 3.3--4 (2005): 171--200.
link.springer.com. Web.

>>>>

In previous work, Billings et. al., discovered through requirements
gathering interviews that many of the difficulties using
high-performance modeling and simulation software fall broadly into five
distinct categories, \cite{billings_designing_2009}. These activities,
detailed in section \ref{workflow-model}, include (1) creating input,
(2) executing jobs, (3) analyzing results, (4) managing data, and (5)
modifying code. There are many tools that address these problems
individually, but the same research found that the excess number and
specialization of these tools also contribute to the learning curve.

Efforts to address these five issues have previously resulted in general
purpose scientific workflow tools like Kepler, 
\cite{ludascher_scientific_2006}, or myopic tools that only satisfy a single
set of requirements for a single piece of software or platform. These
are opposing extremes, but a middle-of-the-road solution is also
possible. A workflow engine could be developed that limits its scope to
high-performance computing (HPC) and to the set of possible workflows
associated with the five previously mentioned activities. With only
minor additional development, a rich application programming interface
(API) could be exposed so that highly-customized solutions could still
be made based on this limited workflow engine.

It is not clear which, if any, of these solutions is better than the
others, and practical requirements will ultimately dictate the path of a
project's progress. This work considers a middle ground solution and
presents the Eclipse Integrated Computational Environment (ICE) as proof
that it is possible to create such a system. Specifically, the work
described here shows that:

\begin{itemize}
\item
  Modeling and simulation activities can be described in a succinct
  workflow model, (see ``Workflow Model'').
\item
  An architecture for such a workflow system can satisfy the model of
  workflows in an extensible way, (see ``Software Architecture'').
\item
  Such a system is applicable to a suite of problems in energy science,
  including virtual battery simulations, additive manufacturing and
  other areas, (see ``Illlustrative Examples'').
\end{itemize}

This section concludes with an introduction of the workflow model addressed by
ICE. Section \ref{software-description} presents the details of the software
from an architecture perspective while section \ref{illustrative-examples}
provides a set of comprehensive examples. Finally, this paper concludes with a
presentation of the impact, (section \ref{impact}), and sample code,
(section \ref{sample-code-tutorials-and-other-resources}).

@@

Computational scientists perform a variety of tasks in modeling and
simulation that can be abstracted into a lightweight theoretical
framework based broadly on five high-level \emph{activities}: (1)
creating input, (2) executing jobs, (3) analyzing results, (4) managing
data, and (5) modifying code. Those same computational scientist would
most likely find these activities difficult for all codes with which
they lack experience, whereas with their own codes---or those with which
they are most familiar---these tasks may be so simple that they are
taken for granted. Any particular combination of these activities across
one or more scientific software package or code results in a unique
\emph{workflow}. Such a workflow is normally, but not always, requested
by a human user and orchestrated by a \emph{Workflow Management System
(WMS)}.

The most obvious workflow for any individual simulation code or
collection of codes is to string the activities together: the user's
workflow is to create the input, launch the job, perform some analysis,
and manage the data---possibly modifying the code in the process. There
are many other combinations however, including re-running jobs with
conditions or modifications or analyzing someone else's data (See
footnote 1).\footnote{The
authors have identified many different combinations of these activities that
define workflow ``classes.'' When possible, every effort is made to give the
classes funny names such as ``The Re-Run'' or ``The Graduate Student.''}

\textbf{Creating input} is the process of describing the physical model
or state of a system that will be simulated. This could include creating
an input file(s) or making calls to an external process to configure a
running program. In most situations, a computational scientist will
modify existing input or create new input from a template. ``Input''
generally includes runtime parameters for the simulation framework
(e.g., tolerances); configuration options (e.g., data locations, output
locations, module configurations); properties of the materials to be
simulated; and a discretization of the simulation space (e.g., mesh,
grid, particle distribution). The collection of all required input can
be quite large and may go by many names, including ``input set,''
``input package,'' ``problem,'' or, simply, ``input.'' Often, the set of
input files will be described in a ``main'' input file that acts as a
kind of manifest, describing and providing links to all necessary
information for a given problem.

In this work, it should be assumed---unless otherwise noted---that
``input'' refers to the entire set of input, not to a single file.

\textbf{Executing jobs}, or ``running the workflow'' in this context, is
the process of performing calculations using a simulation code or
framework based on known variables---the input. Models and simulation
codes are typically run locally for small jobs or for development. Large
simulations, on the other hand, typically require a large amount of
hardware resources; these resources are usually off site (i.e.,
physically unavailable to the user) and are accessed remotely remotely
through Secure Shell (SSH) connections or similar protocols. Remote
execution requires moving the input in advance of the execution and
copying or moving the output to the user's machine. In many cases,
though, the output is too large to move to the user's local machine.

Local and remote jobs are often monitored to ascertain a job's status.
This monitoring may be a simple check as to whether or not the execution
has completed, or it may involve monitoring the output of individual
quantities to examine the calculation state. The latter is often used to
detect calculation errors that will result in incorrect results. If such
problems are found, the job is typically cancelled (``killed'') to save
compute resources and is then re-run later.

In this work, it should be assumed---unless otherwise noted---that
``executing a job'' includes monitoring that job in one or more ways,
possibly including real-time updates to visualizations.

\textbf{Analyzing results} includes executing special jobs to transform
data in one or more prescribed ways and producing artifacts with
scientific significance from the transformed data. This may include, for
example, post-processing results and visualizing the new data with
dedicated visualization tools. For many types of scientific computing,
this includes viewing the results of a simulation on a mesh or grid and
extracting publication quality images or movies from that data. Other
cases may include analyzing results in preparation for follow
simulations or performing feature extraction, classification, or
activities for machine learning and data mining.

While this has many similarities to executing a job, it is distinctly
different because the activity changes focus to satisfy the needs of a
human operator. Simple data reduction, where the exact reduction is
known, certainly qualifies as executing a job; however, analysis of
model and simulation results is far from simple data reduction, and is
generally far more interactive for computational scientists.

\textbf{Managing data} includes moving, copying, storing, sharing, or
otherwise interacting with data for or from simulations. This activity
is the most pervasive because each of the other activities requires
interacting with data in some way. In many cases, though, data is still
managed for its own purposes, without performing a simulation,
generating new input, or analyzing results. Examples include archiving
data, packaging data for publications, and updating values manually
(often in light of new information from publications).

\textbf{Modifying code} is not typically considered a part of a
scientific computing workflow. However, modeling and simulation use
cases often require users to explicitly modify code before execution,
such as with the computational fluid dynamics code Nek5000,
\cite{the_nek5000_team_nek5000_2014}, or to issue special re-build instructions.
Many computational scientists consider ``their workflow'' to be re-running
software after modifications for purely exploratory purposes. This may
be required if the model that the author is manipulating cannot be
configured directly as part of the input but can be easily manipulated
by hand.

@@

This model of workflows differs significantly from those of similar
efforts in workflow science because it defines workflows in terms of
activities. Other workflow models in the literature define a workflow as
a collection of computing processes. For example, Yu and Buyya define
grid workflows as ``a collection of tasks that are processed on
distributed resources in a well-defined order to accomplish a specific
goal,'' \cite{yu_taxonomy_2005}. Others, such as Pizzi et al., subscribe to
similar definitions, \cite{pizzi_aiida:_2016}. This ``process'' view is
acceptable where the workflow is static and does not require additional
human input or ``human in the loop'' behavior after the all the initial
human input is provided. That is, a ``process-oriented'' definition is
acceptable where all human input is provided in advance. However,
workflows within ICE are fully interactive with regular call-backs to
humans. It is simpler to discuss ``activities'' than it is to create a
distinction between ``human processes'' and ``computer processes.''
Focusing on ``activities'' over processes (human or computer) also has
the benefit of removing concrete elements such as hardware properties or
software details that distract from details of workflows and WMSes. That
is, considerations such as memory usage and raw performance are
important, but questions about the abstract workflow or what the WMS
should do are far more important.

@@

Workflows and tasks in ICE are not explicitly treated as trees -
Directed Acyclic Graphs, (DAGS) - as is common with grid workflow tools,
\cite{yu_taxonomy_2005}. Instead, ICE's design is inspired by Representational
State Transfer (REST), \cite{fielding_architectural_2000}.

@@

Individual components of workflows, i.e. - workflow ``tasks'' or
``nodes'' - are either encoded directly in the workflow's subclass of
Item or provided as \emph{Actions} that are dynamically registered with
an \emph{Action Factory} and obtained at runtime. Table 1. describes the
differences between Items, Actions, and Forms.\footnote{An upcoming update to the API will see the formal introduction of
IWorkflow, IWorkflowTask, and IWorkflowEngine interfaces to bring ICE's
API language closer to other systems. This must be done with care to
satisfy backwards compatibility requirements on the present API.}

@@

\begin{table*}[t]
\begin{tabularx}{\textwidth}{|l|X|l|}
\hline
Class & Class Description & Object Description\tabularnewline\hline
Item & Java class with code to execute an abstract workflow. Provides a
Form. & Concrete workflow executor.\tabularnewline\hline
Form & Description and template of the data needed for the Item to
process the workflow. & User-modified workflow data.\tabularnewline\hline
Action & Java class for executing a specific task in the workflow. Used
by Items. & Concrete workflow task executor.\tabularnewline\hline
\end{tabularx}
\caption{Class Descriptions for Items, Forms and Actions}
\end{table*}

@@

ICE handles persistence using a \emph{persistence provider}. The only
default persistence provider uses JAX-RS, \cite{burke_restful_2010}, to write
Forms to XML. In principle other persistence providers could replace the
XML-based provider since it is registered as a dynamic service, and in
the past a JAX-P based provider has been used.

@@

ICE is an Eclipse Rich Client Platform (RCP) application,
\cite{mcaffer_eclipse_2010}. It has a plugin architecture based on Equinox, the
reference implementation of the Open Service Gateway Initiative (OSGi) framework specification, \cite{mcaffer_osgi_2010}. It uses over 1200 additional
packages from the Eclipse ecosystem to provide services such as language
support and visualization to users. Each unique element of ICE described
in this work, including the client, core, all Items, data structures and
many others are provided as plugins to Equinox. Most plugins are
dynamically managed and provided as services that can be obtained from a
service registry. All file I/O in ICE, with only a few exceptions,
interacts with the RCP's Resources plugin. This includes remote
resources that are managed with the Eclipse Parallel Tools Platform,
\cite{tibbitts_integrated_2009}.

@@

There are many situations where graphically configuring workflows is
unacceptable, such as when a very large number of workflows will be
executed or the type of information required is very fine-grained. In
these cases, it is often necessary to provide scripts to the workflow
management system. ICE includes the Eclipse Advanced Scripting
Environment, \cite{pontesegger_eclipse_2015}, (EASE), for scripting because it
provides a native way to script Eclipse RCP projects in Javascript,
Jython, and Python. This also makes it possible to extend the
environment by adding Items in these languages.

@@

Outside of the development team, ICE is commonly deployed as a
sophisticated user environment for computational science projects (see
``Impact'' below) and as a visualization tool. ICE originally a
significant amount of visualization support, but at the request of users
in the community that support was ``spun off'' in early 2016 as a
separate visualization project, the Eclipse Advanced Visualization
Project, EAVP, \cite{billings_eclipse_2015}.

@@

Pannala et al. developed a Virtual Integrated Battery Environment (VIBE)
as part of their research into safety and performance characteristics of
lithium ion batteries, \cite{pannala_multiscale_2015}. VIBE provides ICE to
users as part of its distribution. New workflows were added to ICE to create
multiple types of input, configure the simulation software, and launch
simulations of virtual batteries. Interactive 3D visualizations of the
results were embedded in the launcher so that users would quickly find
their results. Figure 1 shows the results from a temperature
distribution of a prismatic cell battery simulated during discharge.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/vibe_20151016.png}
\caption{ICE workbench for VIBE analysis.}
\end{figure}

@@

The MOOSE Framework is a powerful, easy to use multiphysics framework
developed at the Idaho National Laboratory, \cite{gaston_moose:_2009}. ICE
provides workflow tools for MOOSE as well as specialized class
generation utilities for developing custom MOOSE kernels. Many of the
MOOSE tools in ICE were developed closely with the MOOSE team to closely
reproduce various aspects of MOOSE's user interface, Peacock. Figure 2
shows an example of the ICE workbench for a simple structural mechanics
problem solved with the MOOSE framework, \cite{mccaskey_scientific_2015}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/ice-moose.png}
\caption{ICE workbench for MOOSE workflows.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/ice-bjm.png}
\caption{ICE workbench for binder jet modeling.}
\end{figure}

@@

ICE includes a small utility for simulating neutron reflectivity and
comparing the results to data, \cite{billings_brand_2015}. This was developed
in collaboration with colleagues at the Spallation Neutron Source to
replace an older utility written in Visual Basic and distributed via
Excel macros. The newer utility in ICE is capable of processing a single
workflow and is shown in figure 4.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/reflectivity-screenshot.png}
\caption{ICE workbench for neutron reflectivity.}
\end{figure}

@@

As quantum computing grows, so to does the need for sophisticated
software that can utilize quantum hardware or perform calculations on
simulated quantum hardware. Humble et al. created a simulator for
adiabatic quantum computers where workflows were added to ICE to support
interaction with the simulator, but also to process large sets of
quadratic binary optimimzation problems, \cite{humble_integrated_2014}. Figure 5
shows the workbench for this project.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/jaded.png}
\caption{ICE workbench in JADE for quantum computing simulations.}
\end{figure}

ICE also supports other quantum computing projects, including the
eXtreme-scale ACCelerator (XACC) programming framework,
\cite{mccaskey_ornl-qci/xacc_2016}. XACC is a programming framework for
extreme-scale, post-exascale accelerator architectures that integrates
alongside existing conventional applications.

@@

There are many examples of ICE's role in modeling and simulation
projects for nuclear energy, but readers are referred to the work in
\cite{billings_domain-specific_2015} for an extreme example of the level of
customization that is possible in ICE. Support for the ``Reactor
Analyzer'' was dropped in ICE 2.1.8, but it demonstrated ICE's ability
to integrate many different nuclear energy tools for complex analysis.

@@

One pressing area of interest and impact is that of interoperability
between workflow systems. There is significant prior art in combining
large workflow management systems, such as that Mandal et al.,
\cite{mandal_integrating_2007}, but the end goal of gaining the greatest
advantage by using the best capabilities from multiple systems remains elusive.
ICE's unique perspective on workflows and its well-defined API make it
possible to integrate multiple systems in a straightforward way. This
allows it to connect to other workflow environments, such as Triquetrum,
quite easily, \cite{brooks_introducing_2016}. Triquetrum, (like Kepler in
\cite{mandal_integrating_2007}), is a Ptolemy-based workflow engine,
\cite{brooks_triquetrum:_2015}. Several of the authors of this paper are using
it to investigate these issues.

@@

A new cloud based development tool based on ICE is under development by
RNET LLC out of Dayton, Ohio in response to an SBIR award from the U.S.
Department of Energy. This web based version of ICE will continue ICE's
support for nuclear energy and integrate with cloud services such as
Amazon Web Services and Oak Ridge National Laboratory's Compute and Data
Environment for Science (CADES). Additionally, while ICE has not led to
the development of ``spin-off'' companies, ICE source code has been used
in two spin-off projects: EAVP (mentioned earlier) for advanced
visualizations and the Eclipse January project for scientific data
structures, \cite{graham_eclipse_2016}. ICE was also one of the founding
projects of the Eclipse Science Working Group.

Eclipse ICE is an open source project and the authors welcome and encourage
engagement and contributions from readers. Interested parties my inquire by
contacting the corresponding author or visiting the website.

@@

The quintessential resource for information on ICE is the projects web
page, \cite{billings_eclipse_2016}. The ``Resources'' menu includes links to
detailed tutorials and documentation of multiple types. Examples of how to create
new workflow Items are available at
\url{https://github.com/eclipse/ice/tree/master/org.eclipse.ice.demo}.
Examples of how to use the scripting engine (in this case for neutron
reflectivity) are available at
\url{https://github.com/eclipse/ice/tree/master/examples}. ICE includes an
extensive test suite of unit, integration, and UI tests and these are
also excellent examples of how to work with the platform.

@@

See Table \ref{codeTable}.

\begin{table*}[!htbp]
\begin{tabularx}{\textwidth}{|l|X|X|}
\hline
C1 & Current Code Version & `next'\tabularnewline\hline
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/eclipse/ice/tree/next}
\tabularnewline\hline
C3 & Legal Code License & Eclipse Public License 1.0 \tabularnewline\hline
C4 & Code versioning system use & Git \tabularnewline\hline
C5 & Software code languages, tools, and services used & Java, OSGi, Eclipse Rich Client Platform,
and Maven \tabularnewline\hline
C6 & Compilation requirements, operating environments \& dependencies & Java 1.8 or greater, Maven, and
an internet connection for dependencies \tabularnewline\hline 
C7 & If available Link to developer documentation/manual &
\url{https://wiki.eclipse.org/ICE} \tabularnewline\hline 
C8 & Support email for questions & ice-dev@eclipse.org \tabularnewline\hline
\end{tabularx}
\caption{Current Code Version}
\label{codeTable}
\end{table*}

@@

See Table \ref{execTable}.

\begin{table*}[!htbp]
\begin{tabularx}{\textwidth}{|l|X|X|}
\hline
S1 & Current Code Version & 2.2.1 \tabularnewline\hline
S2 & ermanent link to executables of this version &
\url{https://www.eclipse.org/downloads/download.php?file=/ice/builds/2.2.1/}
 \tabularnewline\hline 
S3 & Legal Software License & Eclipse Public License 1.0 \tabularnewline\hline
S4 & Computing platforms/Operating Systems & Windows (32/64-bit), Mac OS/X,
Linux (32/64-bit) \tabularnewline\hline 
S5 & Installation requirements \& dependencies & Java 1.8 or
greater \tabularnewline\hline
S6 & If available, link to user manual - if formally published include a
reference to the publication in the reference list &
\url{https://wiki.eclipse.org/ICE} \tabularnewline\hline 
S7 & Support email for questions & ice-users@eclipse.org \tabularnewline\hline
\end{tabularx}
\caption{Current Executable Software}
\label{execTable}
\end{table*}
